---
title: "Mock community example"
author: "Christian Diener"
output:
    html_document:
        toc: true
        toc_float: true
        theme: paper
    pdf_document:
        latex_engine: lualatex
        toc: true
        number_sections: true
---


For installtion instructions please see https://github.com/cdiener/microbiome.
We recommend using the docker image in the cloud or locally as this ensures
that all requirements are fulfilled in the correct version.

## Loading dependencies

In order to facilitate use of the pipeline we provide the `mbtools` R package
in this repository which serves two major purposes:

1. It implements additional helper functions for the analysis and benchmarking
   of microbial community data.
2. It depends on all additional packages required for analysis and load them
   upon import.

So loading `mbtools` should be the first step when running the analysis

```{r}
library(mbtools)
```

## Getting the mock community data

`mbtools` includes helper functions to obtain benchmark mock data sets from
the [mockrobiota database](https://peerj.com/preprints/2065/). For instance
to download the mock-3 data set (relatively small) we can use

```{r}
if (!file.exists("mock3.rds")) {
    mock <- mockrobiota("mock-3", "mock3")
    saveRDS(mock, "mock3.rds")
} else mock <- readRDS("mock3.rds")
```

Here `mock` now includes annotations for the data set as a list.

```{r}
names(mock)
mock$samples
```

## Preparing the reads

As we can see we have 3 read files, the forward and backward reads and one
index file mapping the sample barcodes to the sequences. However, we have 4
samples: a uniform and staggered community in duplicates each. In order to
map sequence variants to samples `dada2` expects read files to be separated by
sample. `mbtools` includes helper functions to obtain this splitting.

```{r}
reads <- c(mock$forward, mock$reverse)
barcodes <- mock$samples$BarcodeSequence
names(barcodes) <- mock$samples[,1]
bcs <- split_barcodes(reads, mock$index, "split", barcodes)
fwd <- list.files("split", pattern="forward", full.names=T)
bwd <- list.files("split", pattern="reverse", full.names=T)

fwd
```

As we see that now gives us the reads separated by sample. The orginal read
still include some valid information, particularly they include the read qualities
across all samples.

```{r}
plotQualityProfile(reads[1])
plotQualityProfile(reads[1])
```

As we can see both qualities detoriate extremely with read lengths. Thus,
we will now pass trim the reads. From the plots we can see that the forward
reads have decent quality up to a length of 150 bp whereas the reverse reads
are acceptable up to 100 bp.

```{r}
dir.create("filtered")
fwd_filt <- file.path("filtered", basename(fwd))
bwd_filt <- file.path("filtered", basename(bwd))
for (i in seq_along(fwd)) {
    fastqPairedFilter(c(fwd[i], bwd[i]), c(fwd_filt[i], bwd_filt[i]),
        truncLen=c(150, 100), compress=T, verbose=T)
}
```

We will follow by dereplicating the reads which will yield the unique
sequences in the samples.

```{r}
derepFs <- derepFastq(fwd_filt, verbose=TRUE)
derepRs <- derepFastq(bwd_filt, verbose=TRUE)
# Name the derep-class objects by the sample names
names(derepFs) <- names(derepRs) <- names(barcodes)
```

## Obtaining the sequence variants (sequence OTUs)

With the trimmed and dereplicated reads we can now advance to running the dada2
algorithm to discover the unqiue sequence variants in our reads. We will do this
separately for the forward and backward reads.

```{r}
dadaFs <- dada(derepFs, err=NULL, selfConsist = TRUE)
dadaRs <- dada(derepRs, err=NULL, selfConsist = TRUE)
```

This will fit an error model that deconvolutes the original sequence variants
in the sample. We can investigate how well the error model reproduces our data
as well.

```{r}
plotErrors(dadaFs, nominalQ=TRUE)
plotErrors(dadaRs, nominalQ=TRUE)
```

We will now quantify the sequence variants for both samples and combine them.
Normally we would try to actually combine the forward and backwards reads into
larger reads and quantify those, however our read qualities were so bad in this
data set that there is no sufficient overlap. Thus, we will treat the forward and
backward reads independently.

```{r}
seqtab <- cbind(
    makeSequenceTable(dadaFs),
    makeSequenceTable(dadaRs))
```

Finally, we will also remove bimeras from the data set.

```{r}
seqtab_nochim <- removeBimeraDenovo(seqtab, verbose=TRUE)
```

## Taxonomy assignment and validation

In order to classify taxonomies for the individual sequence variants we will
use the RDP algorithm with the green genes reference sequences. If you do not
use the docker image this data set has to be downloaded first [from here](https://www.dropbox.com/sh/mfcivbudmc21cqt/AAB1l-AUM5uKvjrR33ct-cTXa).

```{r}
file.exists("gg_13_8_train_set_97.fa.gz")
```

Axonomies can now be assigned by

```{r}
taxa <- assignTaxonomy(seqtab_nochim, "gg_13_8_train_set_97.fa.gz")
colnames(taxa) <- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")
head(unname(taxa))
```

This can be combined with quantifications using the phyloseq package.

```{r}
ps <- phyloseq(otu_table(seqtab_nochim, taxa_are_rows=FALSE),
               tax_table(taxa))
```

We can create a table quantifying the taxa across samples with

```{r}
taxa <- cbind(as.data.frame(tax_table(ps)), t(otu_table(ps)))
rownames(taxa) <- NULL

head(taxa)
```

We can now compare this table to our reference table from mockrobiota. We will
start by quantifying how many of the real taxa were found in our taxonomy
assignment.

```{r}
taxa_metrics(taxa[, 1:7], mock$tax_gg[, 1:7])
```

As we can see we are pretty good in identifying the taxa in our samples.
However, how do we perform in terms of taxa quantification? We will
start by stratifying across the samples:

```{r}
res <- NULL

# ierate over samples
for (i in 8:11) {
    tq <- taxa_quants(taxa[, c(1:7, i)], mock$tax_gg[, c(1:7, i)], normalize=T)
    res <- rbind(res, cbind(tq, sample=colnames(taxa)[i]))
}

head(res)
```

`taxa_quants` already qunatifies across a possible levels of taxonomy. So we
can plot the performance stratified by sample with

```{r}
ggplot(res, aes(x=reference, y=measured, col=level)) +
    geom_abline(alpha=0.5) +
    geom_point() +
    facet_wrap(~ sample)
```

Additionally, we can also combine all samples and rather stratify by taxonomy
level.

```{r}
ggplot(res, aes(x=reference, y=measured, col=sample)) +
    geom_abline(alpha=0.5) +
    geom_point() +
    facet_wrap(~ level, scale="free")
```

This looks pretty ok with larger variations on the species level, which is to
be expected. We can quantify the performance by a correlation test.

```{r}
cor.test(res$measured, res$reference)
```

So we get a correlation of about 0.85 which is okay for the low number of reads
we used here.
